DCMIP2016, Part 1: Models and Equation Sets

### 2016-08-18 contributions/ICON/ICON-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

\Author[6]{Thomas}{Dubos}
\Author[7]{Marco}{Giorgetta}
\Author[8]{Elijah}{Goodfriend}
\Author[9]{David A.}{Hall}
\Author[10]{Lucas}{Harris}
\Author[8]{Hans}{Johansen}
\Author[11]{Christian}{Kuehnlein}
\Author[12]{Vivian}{Lee}
\Author[13]{Thomas}{Melvin}
\Author[14]{Hiroaki}{Miura}
\Author[15]{David}{Randall}
\Author[16]{Alex}{Reinecke}
\Author[4]{William}{Skamarock}
\Author[16]{Kevin}{Viner}
\Author[17]{Robert}{Walko}

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{Institut Pierre-Simon Laplace (IPSL)}
\affil[7]{Max Planck Institute for Meteorology}
\affil[8]{Lawrence Berkeley National Laboratory}
\affil[9]{University of Colorado, Boulder}
\affil[10]{Geophysical Fluid Dynamics Laboratory}
\affil[11]{European Center for Medium-Range Weather Forecasting}
\affil[12]{Environment Canada}
\affil[13]{U.K. Met Office}
\affil[14]{University of Tokyo}
\affil[15]{Colorado State University}
\affil[16]{Naval Research Laboratory}
\affil[17]{University of Miami}

@@

@@

@@

@@

@@

@@

@@

HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model. It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed. Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth. It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom. A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$. The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid. After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing. Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}. Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form. The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios. Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}. The dynamics are also discretized using one of several explicit Runge-Kutta times schemes. Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.

In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model. This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}. In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical. Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$. Fields are current co-located in the vertical and explicit time-stepping is used. Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering. This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability. Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.

@@

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{DCMIP2016.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

>>>>

\Author[6]{Thomas}{Dubos}
\Author[7]{Marco}{Giorgetta}
\Author[8]{Elijah}{Goodfriend}
\Author[9]{David A.}{Hall}
\Author[10]{Lucas}{Harris}
\Author[8]{Hans}{Johansen}
\Author[11]{Christian}{Kuehnlein}
\Author[12]{Vivian}{Lee}
\Author[13]{Thomas}{Melvin}
\Author[14]{Hiroaki}{Miura}
\Author[15]{David}{Randall}
\Author[16]{Alex}{Reinecke}
\Author[18]{Daniel}{Reinert}
\Author[4]{William}{Skamarock}
\Author[16]{Kevin}{Viner}
\Author[17]{Robert}{Walko}

\affil[1]{University of California, Davis}
\affil[2]{University of Michigan}
\affil[3]{University of South Wales}
\affil[4]{National Center for Atmospheric Research}
\affil[5]{Stony Brook University}
\affil[6]{Institut Pierre-Simon Laplace (IPSL)}
\affil[7]{Max Planck Institute for Meteorology}
\affil[8]{Lawrence Berkeley National Laboratory}
\affil[9]{University of Colorado, Boulder}
\affil[10]{Geophysical Fluid Dynamics Laboratory}
\affil[11]{European Center for Medium-Range Weather Forecasting}
\affil[12]{Environment Canada}
\affil[13]{U.K. Met Office}
\affil[14]{University of Tokyo}
\affil[15]{Colorado State University}
\affil[16]{Naval Research Laboratory}
\affil[17]{University of Miami}
\affil[18]{Deutscher Wetterdienst}

@@

The ICON model uses a triangular grid, which for the common global applications is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices. Each vertex shares 5 triangles. These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained. For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle. By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes. The ICON grid generator allows an "arbitrary" subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.

For practical purposes the icosahedron is oriented such that one of the major axes coincides with the Earth's rotation axis, with one arc at the 0 meridian. The mean resolution on a global grid is defined as $dx_{mean} = SQRT(4 \cdot \pi \cdot a_ref^2 / n_c)$. For a discussion of the effective resolution see Dipankar et al. (2016). As a specialty the ICON grid generator can produce additive regional grids. These are regional grids produced by additional refinement steps limited to a region of choice on a given parent grid. If needed several regions can be added to a base grid, and each region can be refined by one or more steps. During a simulation these regions are coupled by one-way or two-way nesting to the global simulation. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

@@

\subsection{ICON} \label{sec:ICONEquations}

The equation system of the ICON model is based upon the prognostic variables suggested by \cite{gassmann2008} 
but uses the two-dimensional rather than the three-dimensional Lamb transformation to convert the nonlinear momentum
advection into a vector-invariant form. So far, the shallow-atmosphere approximation is applied, but there
are plans to add an option for the unapproximated deep-atmosphere equations. The basic equation system reads as follows
\begin{align}\label{1eqmo}
\pdiff{v_n}{t} + \pdiff{K_h}{n} + (\zeta + f)v_t + w\pdiff{v_n}{z} = -c_{pd} \theta_v \pdiff{ \pi}{n} + F(v_n)
\end{align}
\begin{align}\label{3eqmo}
\pdiff{w}{t} + {\bf v_h} \cdot \nabla w + w \pdiff{w}{z} = -c_{pd} \theta_v \pdiff{\pi}{z} - g  
\end{align}
\begin{align}\label{conteq}
\pdiff{\rho}{t} + \nabla \cdot ({\bf v} \rho) = 0
\end{align}
\begin{align}\label{traceq}
\pdiff{\rho q}{t} + \nabla \cdot ({\bf v} \rho q) = S(\rho q)
\end{align}
\begin{align}\label{thdeq1}
\pdiff{\rho \theta_v}{t} + \nabla \cdot ({\bf v} \rho \theta_v) = \tilde{Q}
\end{align}
\rrv{As in \cite{gassmann2013},} Eq. (\ref{thdeq1}) is reformulated using the time derivative of
\begin{align}\label{def-exner}
\pi = \left( \frac{R_d}{p_{00}} \rho \theta_v \right)^\frac{R_d}{c_{vd}}
\end{align}
in order to obtain
\begin{align}\label{thdeq}
\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot ({\bf v} \rho \theta_v) = \hat{Q} \,,
\end{align}
\rrv{which simplifies the implicit numerical treatment of the terms representing vertical sound wave propagation.}
The prognostic variables of the (untransformed) equation system are the horizontal velocity component normal to the triangle edges $v_n$, the
vertical wind component $w$, density $\rho$, and virtual potential temperature $\theta_v$. When coupled with moisture physics,
$\rho$ represents the full air density including liquid and solid hydrometeors. The reconstructed
tangential velocity component is denoted as $v_t$, and in accordance with the model code, it is assumed here that 
$(v_t,v_n,w)$ form a right-handed system. $\bf v_h$ and $\bf v$ denote the horizontal and three-dimensional wind vector,
respectively. The remaining symbols are the vertical vorticity component $\zeta$, 
the Coriolis parameter $f$, the horizontal part of the kinetic energy $K_h = \frac{1}{2} (v_n^2 + v_t^2)$,
the Exner function $\pi$, the specific heat capacities of dry air at constant pressure $c_{pd}$ and volume $c_{vd}$,
the gas constant of dry air $R_d = c_{pd}-c_{vd}$, the gravitational acceleration $g$, \rv{and the reference pressure
$p_{00}$ = 1000 hPa conventionally used for defining the potential temperature.} $\pdiff{}{n}$ denotes a horizontal derivative
in edge-normal direction, i.e. between the two adjacent mass points. Moreover, $F(v_n)$ denotes source
terms for horizontal momentum (to be specified below), and $\tilde{Q}$ and $\hat{Q}$ denote appropriately 
formulated diabatic heat source terms. In the tracer equation $q$ denotes the mixing ration with respect to the full air density $\rho$, and $S(\rho q)$ represents the source term for a tracer $q$.

@@

\subsubsection{ICON}

The vertical discretization is formulated in a generalized height-based system, employing the Lorenz staggering. The default implementation uses the generalized smooth-level vertical coordinate described by  \cite{leuenberger2010}, but also the \cite{galchen1975} formulation is available.

@@

\subsection{ICON}

The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure. Details are described in sections 2.4 and 2.5 of \cite{zangl2014}. Here a brief summary:

\emph{Damping: }In the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps only without incurring numerical stability problems under extreme conditions.
\begin{align}
F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
\end{align}
$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.

Another artificial damping term is Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top. The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.

\emph{Diffusion: }The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$.
\begin{align} \label{smagn2}
F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
\end{align}
\begin{align}
F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
\end{align}
\begin{align}
F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
\end{align}
An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.

A fourth-order computational diffusion is also available for vertical wind speed $w$. This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures. Consequently diffusion of $w$ is employed for the DCMIP SCELL test.
\begin{align}
F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
\end{align}

@@

\subsubsection{ICON}

The time stepping scheme of ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves. No time splitting is used with respect to sound waves, because the ration of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one. Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand. Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied. Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.

@@

\subsection{ICON}

The ICON model discretizes the compressible equations for a shallow atmosphere in vector invariant form for the horizontal wind on a triangular Arakawa C-grid and a smoothed terrain following height based Lorenz grid. The prognostic variables are the normal wind $v_n$ at the edge mid points of full levels, the vertical wind $w$ in the circumcenters of the triangles on half levels and virtual potential temperature $\theta_v$, full air density $\rho$, which includes moisture and hydrometeor densities and tracer mixing ratios $q_x$ with respect to the full air density. The discretization in time employs a two time level predictor corrector scheme, which is explicit in all terms except for those describing the vertical propagation of sound waves. Time splitting is applied between the dynamics that is forced by slow physics on the one hand and horizontal diffusion, tracer transport, and fast physics. One complete time step typically includes 5 dynamical sub-steps. The average air mass flux of the dynamical  sub-steps is provided to the tracer transport to allow for a mass-consistent transport. For stabilization of the divergence term on the triangular C-grid the divergence in a triangle is computed from modified normal wind components resulting from a weighted average including normal winds on edges of adjacent cells. Further divergence damping is applied to the normal wind at every sub-step. Rayleigh damping is applied to the vertical wind in layers close to the model top in order to avoid the reflection of gravity waves. The horizontal diffusion, which is applied at full model time steps, combines a flow dependent Smagorinski scheme with a background 4th order Laplacian diffusion operator. For tracer transport a flux form semi-Lagrangian scheme with monotone flux limiters is used, which grants local mass conservation and consistency with the air motion. The numerical methods have been chosen for high numerical efficiency, and they rely on next neighbour communication only, thus allowing massive parallelization.

@@

@@

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{./DCMIP2016.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}
